Towards atomistic resolution structure of phosphatidylcholine glycerol backbone and choline headgroup at different ambient conditions

### 2015-01-31 HGmodelMANUSCRIPT/HGmodel_draft3.tex
Updated the dehydration plot.
Added modified cholestol figure, acknowledgements of MSM, and fixed typos in addresses.

\author{Alexandru Botan}
\thanks{The authors are listed in alphaphetical order.}
\thanks{The author list is not completed.}
\thanks{Institut Lumi\`ere Mati\`ere, UMR5306 Universit\'e Lyon 1-CNRS, Universit\'e de Lyon 69622 Villeurbanne, France}
%
\author{Andrea Catte}
\thanks{University of East Anglia, Norwich, United Kingdom}
\author{Fernando Favela}
\author{Patrick Fuchs}
\thanks{Institut Jacques Monod, CNRS, Université Paris Diderot, Sorbonne Paris Cité, Paris, France}
%\thanks{Université Paris Diderot, Sorbonne Paris Cité, Paris, France}
\author{Matti Javanainen}
\thanks{Tampere University of Technology, Tampere, Finland}
\author{Waldemar Kulig}
\thanks{Tampere University of Technology, Tampere, Finland}
\author{Antti Lamberg}
\thanks{Department of Chemical Engineering, Kyoto University, Kyoto, Japan}
%\author{Alexander Lyubartsev}
\author{Markus S. Miettinen}
\thanks{Fachbereich Physik, Freie Universität Berlin, Berlin, Germany}
\author{Luca Monticelli}
\thanks{5IBCP, CNRS UMR 5086, Lyon, France}
\author{Jukka M\"a\"att\"a}
\thanks{Aalto University, Espoo, Finland}
%\setcounter{page}{1}
\author{Vasily S. Oganesyan}
\thanks{University of East Anglia, Norwich, United Kingdom}
\author{O. H. Samuli Ollila} 
\thanks{{\bf Author to whom correspondence may be addressed. E-mail: samuli.ollila@aalto.fi.}}
\thanks{Helsinki Biophysics and Biomembrane Group, Department of Biomedical Engineering and Computational Science, Aalto University, Espoo, Finland}
\author{Marius Retegan}
\thanks{Max Planck Institute for Chemical Energy Conversion, Mülheim an der Ruhr, Germany}
\author{Hubert Santuz}
\thanks{INSERM, U1134, DSIMB; Institut National de la Transfusion Sanguine (INTS); Laboratoire d'Excellence GR-Ex, Paris, France}
\thanks{Université Paris Diderot, Sorbonne Paris Cité, Paris, France}
\author{Joona Tynkkynen}
\thanks{Tampere University of Technology, Tampere, Finland}
%\author{Mark Wilson}
%\author{Alexander Vogel}

@@

\subsubsection{Phospholipid bilayer with low hydration level}
The experimental order parameters available in the literature~\cite{dvinskikh05b,ulrich94,bechinger91} 
for the glycerol backbone and choline as a function of hydration level are shown in Fig.~\ref{ordPhydr}. 
The independently reported values for choline segments are in good agreement with each other (despite 
slight differences in temperature and acyl chain composition),
showing increase for both segments with decreasing hydration level. It should be noted that only 
absolute values were measured in the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91}, but
we have included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the $\beta$ order parameter with negative sign actually increases with dehydration 
since the absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
Slight decrease for the glycerol backbone g$_3$- and g$_2$- order parameters were observed with dehydration, 
while g$_1$ remained practically unchanged~\cite{dvinskikh05b}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{OrderParameterDEHYDpresentation.eps}
\todo{Several people think that this plot should be clarified.}
  \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
  }
\end{figure}

@@

The choline headgroup and glycerol backbone order parameters for POPC measured by $^{13}$C NMR~\cite{ferreira13} and DPPC choline order parameters 
measured by $^{2}$H NMR~\cite{brown78} are shown in Fig.~\ref{ordPchol} as a function of cholesterol content.
The agreement between different experimental results is again very good, showing only very modest changes in 
the choline order parameters as a function of cholesterol content. It should be noted, however, that very small
changes are measurable with high resolution $^{2}$H NMR experiments
and cholesterol causes a measurable increase in the $\beta$ order parameter and a forking in the $\alpha$ order
parameter~\cite{brown78}. These effects are, however, so small that they are barely visible in the scale used in Fig.~\ref{ordPchol}.
Further, the effects of cholesterol on the glycerol backbone order parameters for POPC from $^{13}$C NMR experiment~\cite{ferreira13} 
are in good agreement with the results for the phosphatidylethanolamine (PE) measured by $^{2}$H NMR~\cite{ghosh82}.
These results further support the idea that the glycerol backbone structural behaviour is independent of the
headgroup composition~\cite{gally81} and that the headgroup stucture is independent of the acyl chain region content unless
charges are present~\cite{scherer87}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{OrderParameterCHOLexpVSsim.eps}
  \caption{\label{ordPchol}
    The effect of cholesterol content on the glycerol backbone and choline order parameters in experiments~\cite{brown78,ferreira13} and simulations
    with the Berger-POPC-07/H\"oltje-CHOL-13, CHARMM36 and MacRog force fields. The signs in the experimental values are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.  Most order parameters from Berger-POPC-07/H\"oltje-CHOL-13 model for g$_1$ are beoynd the y-axis scale.}
\end{figure}

@@

\section{Conclusions}
The atomistic resolution structures sampled by the glycerol backbone and choline headgroup
in phoshatidylcholine bilayers are not known despite of vast amount of accurate experimental
data. Atomistic resolution molecular dynamics simulation model which reproduced the
experimental data would automatically resolve the structures giving an interpretation of experimental results.
In this work we have collected and reviewed the experimental C--H bond vector order
parameters available in the literature. These experimental parameters are then compared to
different atomistic resolution simulation models for a fully hydrated bilayer, bilayers dehydrated to different extents, and
lipid bilayers containing various amounts of cholesterol. Our results have led to the following conclusions:

@@

\bibliographystyle{apsrev}
\bibliography{refs}

>>>>

\author{Alexandru Botan}
\thanks{The authors are listed in alphaphetical order.}
\thanks{The author list is not completed.}
\thanks{Institut Lumi\`ere Mati\`ere, UMR5306 Universit\'e Lyon 1-CNRS, Universit\'e de Lyon 69622 Villeurbanne, France}
%
\author{Andrea Catte}
\thanks{University of East Anglia, Norwich, United Kingdom}
\author{Fernando Favela}
\author{Patrick Fuchs}
\thanks{Institut Jacques Monod, CNRS, Universit\'e Paris Diderot, Sorbonne Paris Cit\'e, Paris, France}
%\thanks{Université Paris Diderot, Sorbonne Paris Cité, Paris, France}
\author{Matti Javanainen}
\thanks{Tampere University of Technology, Tampere, Finland}
\author{Waldemar Kulig}
\thanks{Tampere University of Technology, Tampere, Finland}
\author{Antti Lamberg}
\thanks{Department of Chemical Engineering, Kyoto University, Kyoto, Japan}
%\author{Alexander Lyubartsev}
\author{Markus S. Miettinen}
\thanks{Fachbereich Physik, Freie Universit\"at Berlin, Berlin, Germany}
\author{Luca Monticelli}
\thanks{5IBCP, CNRS UMR 5086, Lyon, France}
\author{Jukka M\"a\"att\"a}
\thanks{Aalto University, Espoo, Finland}
%\setcounter{page}{1}
\author{Vasily S. Oganesyan}
\thanks{University of East Anglia, Norwich, United Kingdom}
\author{O. H. Samuli Ollila} 
\thanks{{\bf Author to whom correspondence may be addressed. E-mail: samuli.ollila@aalto.fi.}}
\thanks{Helsinki Biophysics and Biomembrane Group, Department of Biomedical Engineering and Computational Science, Aalto University, Espoo, Finland}
\author{Marius Retegan}
\thanks{Max Planck Institute for Chemical Energy Conversion, M\"ulheim an der Ruhr, Germany}
\author{Hubert Santuz}
\thanks{INSERM, U1134, DSIMB; Institut National de la Transfusion Sanguine (INTS); Laboratoire d'Excellence GR-Ex, Paris, France}
\thanks{Universit\'e Paris Diderot, Sorbonne Paris Cit\'e, Paris, France}
\author{Joona Tynkkynen}
\thanks{Tampere University of Technology, Tampere, Finland}
%\author{Mark Wilson}
%\author{Alexander Vogel}

@@

\subsubsection{Phospholipid bilayer with low hydration level}
The experimental order parameters available in the literature~\cite{dvinskikh05b,ulrich94,bechinger91} 
for the glycerol backbone and choline as a function of hydration level are shown in Fig.~\ref{ordPhydr}. 
The independently reported values for choline segments are in good agreement with each other (despite 
slight differences in temperature and acyl chain composition),
showing increase for both segments with decreasing hydration level. It should be noted that only 
absolute values were measured in the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91}, but
we have included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the $\beta$ order parameter with negative sign actually increases with dehydration 
since the absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
Slight decrease for the glycerol backbone g$_3$- and g$_2$- order parameters were observed with dehydration, 
while g$_1$ remained practically unchanged~\cite{dvinskikh05b}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/dehydration.pdf}
\todo{Several people think that this plot should be clarified. ------> DONE.}
  \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
  }
\end{figure}

@@

The choline headgroup and glycerol backbone order parameters for POPC measured by $^{13}$C NMR~\cite{ferreira13} and DPPC choline order parameters 
measured by $^{2}$H NMR~\cite{brown78} are shown in Fig.~\ref{ordPchol} as a function of cholesterol content.
The agreement between different experimental results is again very good, showing only very modest changes in 
the choline order parameters as a function of cholesterol content. It should be noted, however, that very small
changes are measurable with high resolution $^{2}$H NMR experiments
and cholesterol causes a measurable increase in the $\beta$ order parameter and a forking in the $\alpha$ order
parameter~\cite{brown78}. These effects are, however, so small that they are barely visible in the scale used in Fig.~\ref{ordPchol}.
Further, the effects of cholesterol on the glycerol backbone order parameters for POPC from $^{13}$C NMR experiment~\cite{ferreira13} 
are in good agreement with the results for the phosphatidylethanolamine (PE) measured by $^{2}$H NMR~\cite{ghosh82}.
These results further support the idea that the glycerol backbone structural behaviour is independent of the
headgroup composition~\cite{gally81} and that the headgroup stucture is independent of the acyl chain region content unless
charges are present~\cite{scherer87}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/cholesterolization.pdf}
  \caption{\label{ordPchol}
    The effect of cholesterol content on the glycerol backbone and choline order parameters in experiments~\cite{brown78,ferreira13} and simulations
    with the Berger-POPC-07/H\"oltje-CHOL-13, CHARMM36 and MacRog force fields. The signs in the experimental values are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.  Most order parameters from Berger-POPC-07/H\"oltje-CHOL-13 model for g$_1$ are beoynd the y-axis scale.}
\end{figure}

@@

\pagebreak
\section{Conclusions}
The atomistic resolution structures sampled by the glycerol backbone and choline headgroup
in phoshatidylcholine bilayers are not known despite of vast amount of accurate experimental
data. Atomistic resolution molecular dynamics simulation model which reproduced the
experimental data would automatically resolve the structures giving an interpretation of experimental results.
In this work we have collected and reviewed the experimental C--H bond vector order
parameters available in the literature. These experimental parameters are then compared to
different atomistic resolution simulation models for a fully hydrated bilayer, bilayers dehydrated to different extents, and
lipid bilayers containing various amounts of cholesterol. Our results have led to the following conclusions:

@@

MSM acknowledges  financial support from the Volkswagen Foundation (86110).
\bibliographystyle{apsrev}
\bibliography{refs}
